Large and Infinitary Quotient Inductive-Inductive Types

### 2019-12-26 draft/paper.tex
work on paper

\documentclass{llncs}

\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathpartir}
\usepackage{scalerel}

%% Definition env with non-italic body text
\spnewtheorem{nidefinition}[definition]{Definition}{\bfseries}{\textnormal}

%% abbrevs
%% ------------------------------------------------------------

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\mathsf{\Pi}}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\mathsf{\hat{\Pi}}}
\newcommand{\appe}{\mathsf{ap\hat{p}}}
\newcommand{\lame}{\mathsf{la\hat{m}}}
\newcommand{\Piinf}{\mathsf{\tilde{\Pi}}}
\newcommand{\appinf}{\mathsf{ap\tilde{p}}}
\newcommand{\laminf}{\mathsf{la\tilde{m}}}

\newcommand{\appitt}{\mathop{{\scriptstyle @}}}

\newcommand{\Refl}{\mathsf{Refl}}

\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}

\newcommand{\Sg}{\mathsf{\Sigma}}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}

%% ------------------------------------------------------------

@@

\title{Generalizations of Quotient Inductive-Inductive Types\thanks{This work
    was supported by EFOP-3.6.3-VEKOP-16-2017-00002 grant and COST
    Action EUTypes CA15123.}}

\author{Ambrus Kaposi \and Andr{\'a}s Kov{\'a}cs}
\institute{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary \\ \{akaposi$|$kovacsandras\}@inf.elte.hu}

\maketitle

\begin{abstract}
Quotient inductive-inductive types (QIITs) are generalized inductive types which
allow sorts to be indexed over previously declared sorts, and allow usage of
equality constructors. QIITs are especially useful for algebraic descriptions of
type theories and constructive definitions of real, ordinal and surreal
numbers. We develop new metatheory for large QIITs, large elimination, recursive
equations, infinitary constructors and equations between sorts.
%% First, we develop tools which allow precise and convenient handling
%% of universe levels, working internally in an extensional type theory with
%% cumulative universes.
As in prior work, we describe QIITs using a type theory where each context
represents a QIIT signature. However, in our case the theory of signatures can
also describe its own signature. We use self-description to bootstrap a model
theory for the theory of signatures without using preterms or assuming a
pre-existing internal syntax for a type theory. We give initial algebra
semantics for described QIITs, and we show the equivalence of initiality and
induction. We present two extensions of a previous term model construction. The
first one constructs all large infinitary QIITs without sort equations from the
QIIT of the theory of signatures. The second one constructs all large finitary
QIITs with sort equations, from the same syntax. This separation is required
because handling infinitary constructors in the term model requires showing a
strong form of invariance under algebra isomorphisms, which is violated by sort
equations.

\end{abstract}

\section{Introduction}
\label{sec:intro}

ISSUE. We actually need flCwF pseudomorphisms for Sub in infinitary semantics.
$a : \Tm\,\Gamma\,\U$ is \emph{not} a discrete displayed flCwF, only $\El\,a$ is!
$\Piinf\,A\,b : \Tm\,\Gamma,\U$ does not preserve (displayed) object in $\Gamma$
strictly.

\begin{itemize}
\item New semantics with pseudomorphisms for Sub and Tm. Is it enough to
      have weak preservation of comprehension, and we get all other preservation
      for free, see ``gluing for type theory''?
\item Check term model and eliminators to see if change needed.
\item Check if the isomorphism for K in flCwF can be made strict, i.e.
      $\Sub\,\Gamma\,\Delta = \Tm\,\Gamma\,(\Kfam\,\Delta)$.
\end{itemize}

@@

Categories are a good example. Signatures for QIITs allow having multiple sorts,
with later ones indexed over previous ones, and equations as well. We need both
features in order to write down the signature of categories:
\begin{alignat*}{3}
  & Obj && : \Set \\
  & Mor && : Obj \ra Obj \ra \Set \\
  & id  && : Mor\,i\,j \\
  & \blank\circ\blank && : Mor\,j\,k\ra Mor\,i\,j \ra Mor\,i\,k \\
  & idl && : id \circ f = f \\
  & idr && : f \circ id = f \\
  & ass && : f \circ (g \circ h) = (f \circ g) \circ h
\end{alignat*}
The benefit of having a QIIT signature is getting a model theory ``for free'',
from the metatheory of QIITs. This model theory includes a category of algebras
which has an initial object and also some additional structure. For the
signature of categories, we get the empty category as the initial object, but it
is common to consider categories with more structure, which have more
interesting initial models.

Algebraic notions of models of type theories are examples for this. Here,
initial models represent syntax, and initiality corresponds to induction on
syntax. A number of different notions have been used, from contextual categories
and comprehension categories to categories with families, but all of these are
categories with extra structure.

The main motivation of the current paper is to extend the notion of QIIT so that
it accommodates all algebraic notions of type theories which have been used in
previous works. As a side effect of fulfilling this goal, infinitary QIITs such
as Cauchy real numbers are covered as well.

We generalize previous notions of QIITs in the following ways:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and models at different
  universe levels. This feature is routinely used in the metatheory of type
  theory, but it has not been presented explicitly in previous works about
  QIITs.
  \item
  \textbf{Infinitary constructors}. This allows infinitely branching
  trees. Also, the theory of QIIT signatures is itself large and infinitary,
  thus now it can ``eat itself'', i.e.\ include its own signature and provide
  its own metatheory. This was previously not possible in \cite{TODO}, where
  only finitary QIITs were described. We use this self-representation to
  bootstrap the model theory of signatures, without having to assume any
  pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions\cite{TODO}.
  \item
  \textbf{Sort equations}, or equations between type constructors. They have
  been used recently to conveniently represent type-theoretic universes, for
  example Russell-style universes\cite{TODO} or cumulative hierarchies\cite{TODO}.
\end{enumerate}

We also develop semantics. We show that for each signature, there is a CwF
(category with families) of algebras, extended with $\Sigma$-types, extensional
identity, and constant families. This additional structure corresponds to a
type-theoretic flavor of finite limits, and it was shown in \cite{TODO} that the
category of such CwFs is biequivalent to the category of finitely complete
categories.

As to the existence of initial algebras, we present two different term model
constructions, yielding the following results:

\begin{enumerate}
\item
  All large infinitary QIITs with recursive equations, but \emph{without
  sort equations}, are reducible to the theory of signatures, i.e.\
  their initial algebras are constructible from the initial algebra for
  the theory of signatures.
\item
  All large finitary QIITs with sort equations, but no recursive equations,
  are reducible to the theory of signatures.
\end{enumerate}

The reason for the two separate constructions is that sort equations are not as
well-behaved as the other extensions: they are modelled as strict equalities of
sets in algebras, hence they do not respect isomorphism of sets. We make
essential use of invariance under set isomorphism in the first term model
construction, so we cannot throw sort equations into that mix.

In Section \ref{sec:levels} we set up cumulative universes which we use in later
sections. In Section \ref{sec:tos} we introduce the theory of QIIT
signatures. In Section \ref{sec:categorical_semantics} we develop initial algebra semantics,
also covering the theory of signatures itself. In Sections
\ref{sec:infinitary_term_model}-\ref{sec:sort_term} we present the two term model
constructions. We discuss related work and conclusions in Sections
\ref{sec:related}-\ref{sec:conclusion}.

\section{Cumulative Extensional Type Theory}
\label{sec:levels}

In the following sections, we will consider QIIT algebras at arbitrary finite
levels, along with large eliminations, where the initial algebra can be at a
different level than the target algebra. For a simple example, consider natural
number algebras at level $i$, given as the $\Sigma$-type $NatAlg_i :\equiv(Nat :
\Set_i)\times Nat\times(Nat \ra Nat)$. The initial such algebra is the set of
natural numbers, which is at level $0$, but in type theory we often want to
eliminate into larger $\Set_i$, for example when computing a $Nat$-indexed
family of types. To support convenient reasoning about levels in this paper, we
need two features: cumulativity and a way to quantify over finite levels.

%% Universe levels are usually viewed as a bureaucratic and tedious part of type
%% theory, and are often omitted or mentioned only in passing. Here we intend to
%% model universe levels in a precise way, because large eliminations are an
%% essential part of the practical usage of inductive types. Hence, we need a way
%% to conveniently reason about levels without too much administrative
%% burden. Unfortunately, simple universe setups are insufficient for this
%% purpose. We need two features: cumulativity and a way to quantify over finite
%% levels.

@@

We need cumulativity to reduce bureaucratic overhead. In its most basic form,
cumulativity requires that whenever $A : \Set_i$, and $i < j$ then also $A :
\Set_j$. We also need cumulativity for $\Sigma$ and function types, in order to
have cumulativity for algebras in general. For example, we want to have $\gamma
: NatAlg_j$ whenever $\gamma : NatAlg_i$ and $i < j$. We have found that
non-cumulative algebras induce excessive ``lifting'' noise, which we would like
to avoid.

%% However, the $NatAlg$ example above also required a form of
%% cumulativity for $\Sigma$ and $\Pi$ types, which we also need in general for
%% QIIT algebras.

We use Sterling's cumulative algebraic type theory \cite{TODO} as the general
setting for the rest of the paper. The reason for this is twofold. First, it
supports the right kind of cumulativity for our purposes. Second, it is itself a
finitary QIIT with sort equations, hence this paper provides a model theory for
it\footnote{Which is, of course, somewhat circular, but not more circular than
  e.g.\ the study of models of ZFC in ZFC.}.

We extend the base theory with an extensional identity type, $\Sigma$-types and
the unit type. In \cite{TODO} a proof of canonicity is provided for the base
theory, which also includes a standard set-theoretic model. It is
straightforward to extend the canonicity proof to cover our additional type
formers.

@@

We also have a lifting operation on types, which introduces cumulativity.
\vspace{-1.5em}
\begin{figure}
\begin{mathpar}
  \inferrule*[lab=universe formation]
             {i < j}
             {\Gamma \vdash \Set_i : \Set_{j}}

@@

  \inferrule*[lab=type lifting]
             {\Gamma \vdash A : \Set_i \\ i < j}
             {\Gamma \vdash\,\Lift_j A : \Set_{j}}

  \inferrule*[lab=lift composition]{}{\Lift_j (\Lift_i A) =\,\,\Lift_{j} A}

  \inferrule*[lab=term lifting]
             {}
             {\{t\,|\,\Gamma \vdash t : A\} = \{t\,|\,\Gamma \vdash t :\,\,\Lift_i A\}}

  \inferrule*[lab=context lifting]
             {}
             {(\Gamma,\,x : A) = (\Gamma,\,x :\,\,\Lift_i A)}

  \inferrule*[lab=universe lifting]{}{\Lift_j \Set_i = \Set_i}

  \inferrule*[lab=function lifting]
             {}
             {\Lift_i((x : A)\ra B) = (x :\,\Lift_i A)\ra\,\Lift_i B}

  \inferrule*[lab=pair lifting]
             {}
             {\Lift_i((x : A)\times B) = (x :\,\Lift_i A)\,\times\Lift_i B}

  \inferrule*[lab=unit lifting]
             {}
             {\Lift_i\top = \top}
\end{mathpar}

@@

In Figure \ref{lifting}, we include an excerpt of cETT's rules. We change
presentation slightly from Sterling: we use a $\U_{\max(i,j)}$ return type in
functions and pairs, instead of having both $A$ and $B$ types in the same
$\Set_i$. These changed rules are all derivable from the original ones. In
general, we can expect that lifting never impedes constructions, because it
appropriately computes out of the way.

Of special note is the \emph{term lifting rule}. It is a sort equation, an
equation between sets of terms, expressing that lifted types have exactly the
same terms as unlifted ones. This allows us to have $t :\,\,\Lift_j A$ whenever $t
: A$. Together with the universe lifting rule, this implies $A : Set_{i+j+1}$
whenever $A : Set_i$. Similarly, the lifting rules for functions and pairs give
us cumulativity for $NatAlg$. We derive a notion of subtyping:

\begin{nidefinition}[Cumulative subtyping]
  For $A : \Set_i$ and $B : \Set_{i+j+1}$, we define $A \prec B : \Set_{i+j+1}$
  as $B =\,\,\Lift_j A$.  It follows from term lifting and equality reflection
  that whenever $A \prec B$ and $t : A$, then also $t : B$.
\end{nidefinition}

@@

We also need to quantify over universe levels. cETT does not support this, and
we leave it like that. Instead, for the sake of simplicity, we quantify over
levels in an unspecified metatheory \emph{outside} cETT. Hence, a universe
polymorphic cETT term is understood as a $\mathbb{N}$-indexed family of
terms. We reuse the notation of cETT functions for level-polymorphism, e.g.\
as in the following function:
\[
\lambda i.\,\Set_i\ra\Set_i : (i : \mathbb{N})\ra \Set_{i+1}
\]
In this paper, we do not need to internalize level-polymorphic constructions,
so this setup is sufficient.

\section{Signatures}
\label{sec:tos}

In this section, we define signatures for QIITs which support all the mentioned
generalizations. Signatures are given as contexts in a certain type theory,
called the theory of signatures. We shall abbreviate it as ToS. However, ToS
turns out to be a large infinitary QIIT itself, and we would like to define ToS
and a notion of signature without referring to QIITs, only using features
present in cETT. As a first step, we define a notion of model.

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

%% ------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathpartir}
\usepackage{scalerel}

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\hat{\Pi}}
\newcommand{\appe}{\mathsf{ap\hat{p}}}
\newcommand{\lame}{\mathsf{la\hat{m}}}
\newcommand{\Piinf}{\tilde{\Pi}}
\newcommand{\appinf}{\mathsf{ap\tilde{p}}}
\newcommand{\laminf}{\mathsf{la\tilde{m}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\Sg}{\Sigma}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}

\newtheorem{abbreviation}{Abbreviation}
%% ------------------------------------------------------------

@@

%% Title information
\title{Large and Infinitary Quotient Inductive-Inductive Types} %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%% \titlenote{with title note}             %% \titlenote is optional;
%%                                         %% can be repeated if necessary;
%%                                         %% contents suppressed with 'anonymous'
%% \subtitle{Subtitle}                     %% \subtitle is optional
%% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
%%                                         %% can be repeated if necessary;
%%                                         %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

\author{Andr{\'a}s Kov{\'a}cs}
\orcid{0000-0002-6375-9781}
\affiliation{
  \department{Department of Programming Languages and Compilers}
  \institution{E{\"o}tv{\"o}s Lor{\'a}nd University}
  \city{Budapest}
  \country{Hungary}                   %% \country is recommended
}
\email{kovacsandras@inf.elte.hu}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Quotient inductive-inductive types (QIITs) are generalized inductive types which
allow sorts to be indexed over previously declared sorts, and allow usage of
equality constructors. QIITs are especially useful for algebraic descriptions of
type theories and constructive definitions of real, ordinal and surreal
numbers. We develop new metatheory for large QIITs, large elimination, recursive
equations and infinitary constructors. As in prior work, we describe QIITs using
a type theory where each context represents a QIIT signature. However, in our
case the theory of signatures can also describe its own signature, modulo
universe sizes. We bootstrap the model theory of signatures using
self-description and a Church-coded notion of signature, without using
complicated raw syntax or assuming an existing internal QIIT of signatures. We
give semantics to described QIITs by modeling each signature as a finitely
complete CwF (category with families) of algebras. Compared to the case of
finitary QIITs, we additionally need to show invariance under algebra
isomorphisms in the semantics. We do this by modeling signature types as
isofibrations. Finally, we show by a term model construction that every QIIT is
constructible from the syntax of the theory of signatures.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10011740</concept_id>
<concept_desc>Theory of computation~Type theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Type theory}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{inductive types, quotient inductive-inductive types}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

@@

Categories are a good example. Signatures for QIITs allow having multiple sorts,
with later ones indexed over previous ones, and equations as well. We need both
features in order to write down the signature of categories.

The benefit of having a QIIT signature is getting a model theory ``for free'',
from the metatheory of QIITs. This model theory includes a category of algebras
which has an initial object and also some additional structure. For the
signature of categories, we get the empty category as the initial object, but it
is common to consider categories with more structure, which have more
interesting initial models.

Algebraic notions of models of type theories are examples for this. Here,
initial models represent syntax, and initiality corresponds to induction on
syntax. Several variants have been used, from contextual categories \cite{TODO} and
comprehension categories \cite{TODO} to categories with families \cite{TODO}, but all of these are
categories with extra structure.

The main motivation of the current paper is to generalize previously formalized
\cite{kaposi2019constructing} QIITs so that they accommodate more algebraic
formulations of type theories, and more aspects of their practical
metatheory. As a side effect of fulfilling this goal, infinitary QIITs such as
Cauchy real numbers \cite{TODO} are covered as well.

\subsection{Contributions}

We add the following features to QIITs:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and algebras at different
  universe levels. Large elimination is routinely used in the metatheory of type
  theory, but it has not been presented explicitly in previous works about
  QIITs.
  \item
  \textbf{Infinitary constructors}. This allows infinitely branching
  trees. Also, the theory of QIIT signatures is itself large and infinitary,
  thus it can ``eat itself'', i.e.\ include its own signature and provide
  its own metatheory. This was not possible previously in
  \cite{kaposi2019constructing}, where only finitary QIITs were described. We
  exploit self-representation to bootstrap the model theory of signatures,
  without having to assume any pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions\cite{TODO}.
\end{enumerate}

To provide semantics, we show that for each signature, there is a CwF (category
with families) of algebras, extended with $\Sigma$-types, extensional identity,
and constant families. This additional structure corresponds to a type-theoretic
flavor of finite limits, as it was shown in \cite{clairambault2014biequivalence}
that the category of such CwFs is biequivalent to the category of finitely
complete categories.

Compared to the case of finitary QIITs, the addition of infinitary constructors
and recursive equations requires a significant change in semantics: instead of
strict CwF morphisms, we need to consider weak ones, and instead of modeling
types as displayed CwFs, we need to model them as CwF isofibrations. In a
nutshell, the new semantics must be given mutually with a form of a ``structure
identity principle''\cite{TODO}, which says that signature extension respects algebra
isomorphisms.

We also show, by a term model construction, that all QIITs are reducible to the
syntax of signatures. This construction also essentially relies on invariance under
isomorphisms.

\subsection{Outline of the Paper}

In Section \ref{sec:metatheory}, we describe a cumulative variant of extensional type
theory which is used as metatheory in the rest of the paper. In Section \ref{sec:tos},
we introduce the theory of QIIT signatures. TODO

\section{Metatheory}

In this paper we consider QII algebras at arbitrary finite levels, along with
large eliminations, where the initial algebra can be at a different level than
the target algebra. For a simple example, consider natural number algebras at
level $i$, given as the $\Sigma$-type $NatAlg_i :\equiv(Nat : \Set_i)\times
Nat\times(Nat \ra Nat)$. The initial such algebra is the set of natural numbers,
which is at level $0$, but in type theory we often want to eliminate into larger
$\Set_i$, for example when computing a $Nat$-indexed family of types. To support
convenient reasoning about levels in this paper, we need two features:
cumulativity and a way to quantify over finite levels. The former is essential
to reduce bureaucratic overhead, while the latter is required to talk about
arbitrary levels.

@@

In its most basic form, cumulativity requires that whenever $A : \Set_i$, and $i
< j$ then also $A : \Set_j$. We also need cumulativity for $\Sigma$ and function
types, in order to have cumulativity for QII algebras in general. For example,
we want to have $\gamma : NatAlg_j$ whenever $\gamma : NatAlg_i$ and $i < j$.

We use Sterling's cumulative algebraic type theory \cite{sterling2019algebraic} as basis for our metatheory. \todo{REASONS}

%% The reason is that it is the only rigorous
%% algebraic definition of cumulativity that we are aware of, and it also

%% The reason for this is twofold. First, it
%% supports the right kind of cumulativity for our purposes. Second, it is itself a
%% finitary QIIT with sort equations, hence this paper provides a model theory for
%% it

We extend the base theory with an extensional identity type, $\Sigma$-types and
the unit type. In \cite{sterling2019algebraic} a proof of canonicity is provided for the base
theory, which also includes a standard set-theoretic model. It is
straightforward to extend the canonicity proof to cover our additional type
formers.

@@

\begin{figure}
\begin{mathpar}
  \inferrule*[lab=universe formation]
             {i < j}
             {\Gamma \vdash \Set_i : \Set_{j}}

@@

  \inferrule*[lab=equality formation]
             {\Gamma \vdash A : \Set_i \\ \Gamma \vdash t : A \\ \Gamma \vdash u : A}
             {\Gamma \vdash t = u : \Set_i}

  \inferrule*[lab=type lifting]
             {\Gamma \vdash A : \Set_i \\ i \leq j}
             {\Gamma \vdash\,\Lift_j A : \Set_{j}}

  \inferrule*[lab=lift composition]{}{\Lift_j (\Lift_i A) \equiv\,\,\Lift_{j} A}

  \inferrule*[lab=identity lift]{\Gamma \vdash A : \Set_i}{(\Lift_i A) \equiv\,A}

  \inferrule*[lab=term lifting]
             {}
             {\{t\,|\,\Gamma \vdash t : A\} \equiv \{t\,|\,\Gamma \vdash t :\,\,\Lift_i A\}}

  \inferrule*[lab=context lifting]
             {}
             {(\Gamma,\,x : A) \equiv (\Gamma,\,x :\,\,\Lift_i A)}

  \inferrule*[lab=universe lifting]{}{\Lift_j \Set_i \equiv \Set_i}

  \inferrule*[lab=function lifting]
             {}
             {\Lift_i((x : A)\ra B) \equiv (x :\,\Lift_i A)\ra\,\Lift_i B}

  \inferrule*[lab=pair lifting]
             {}
             {\Lift_i((x : A)\times B) \equiv (x :\,\Lift_i A)\,\times\Lift_i B}

  \inferrule*[lab=unit lifting]
             {}
             {\Lift_i\top \equiv \top}

  \inferrule*[lab=equality lifting]
             {}
             {\Lift_i(t = u) \equiv (t = u)}
\end{mathpar}

@@

We also have a lifting operation on types, which introduces cumulativity. In
Figure \ref{lifting}, we include an excerpt of cETT's rules. We change
presentation slightly from Sterling: we use a $\U_{\max(i,j)}$ return type in
functions and pairs, instead of having both $A$ and $B$ types in the same
$\Set_i$, and we define type lifting with $i \leq j$ instead of $i < j$. These
changed rules are all derivable from the original ones. In general, we can
expect that lifting never impedes constructions, because it appropriately
computes out of the way.

Of special note is the \emph{term lifting rule}. It is a sort equation, an
equation between sets of terms, expressing that lifted types have exactly the
same terms as unlifted ones. This allows us to have $t :\,\,\Lift_j A$ whenever $t
: A$. Together with the universe lifting rule, this implies $A : \Set_{i+j}$
whenever $A : \Set_i$. Similarly, the lifting rules for functions and pairs give
us cumulativity for $NatAlg$. We derive a notion of subtyping:

\begin{definition}[Cumulative subtyping]
  For $A : \Set_i$, $i \leq j$ and $B : \Set_{j}$, we define $A \leq B : \Set_{j}$
  as $B =\,\,\Lift_j A$.  It follows from term lifting and equality reflection
  that whenever $A \leq B$ and $t : A$, then also $t : B$.
\end{definition}

@@

For the sake of simplicity, we do not add this to cETT directly. Instead, we
quantify over levels in an unspecified metatheory \emph{outside} cETT. Hence, a
universe polymorphic cETT term is understood as a $\mathbb{N}$-indexed family of
terms. We reuse the notation of cETT functions for universe polymorphism, e.g.\ as
in the following function:
\[
\lambda\,i.\,\Set_i : (i : \mathbb{N})\ra \Set_{i+1}
\]
In this paper, we do not need to internalize level-polymorphic constructions,
so this setup is sufficient.

\section{QIIT Signatures}
\label{sec:tos}

Signatures are given as contexts in a certain type theory, the theory of
signatures. We shall abbreviate it as ToS. However, ToS turns out to be a large
infinitary QIIT itself, and we would like to define ToS and a notion of
signature without referring to QIITs, only using features present in cETT. As a
first step, we define a notion of model. $\Pi$

\begin{definition}[Notion of model for the theory of signatures]\label{def:tos}
For levels $i$ and $j$, there is a cETT type $\ToS_{i,j} : \Set_{\max(i,\,j)+1}$, whose elements are
ToS models (or ToS-algebras). $\ToS_{i,j}$ is an iterated $\Sigma$-type, containing
all of the following components.
\begin{enumerate}
\item
  A category with families (CwF), where all four underlying sets (of objects,
  morphisms, types and terms) are in $\Set_j$. Following notation in
  \cite{kaposi2019constructing}, we denote these respectively as $\Con : \Set_j$, $\Sub : \Con \ra
  \Con \ra \Set_j$, $\Ty : \Con \ra \Set_j$ and $\Tm : (\Gamma : \Con) \ra
  \Ty\,\Gamma \ra \Set_j$. We denote the empty context as $\emptycon : \Con$,
  context extension as $\blank\ext\blank : (\Gamma : \Con)\ra \Ty\,\Gamma \ra
  \Con$.  Substitution on types and terms is written as $\blank[\blank]$, with
  $\id$ and $\blank\circ\blank$ for identity and composition, and $\blank,\blank
  : (\sigma : \Sub\,\Gamma\,\Delta)\ra \Tm\,\Gamma\,A[\sigma]\ra
  \Sub\,\Gamma\,(\Delta\ext A)$ for substitution extension.
\item
  A universe $\U : \Ty\,\Gamma$ with decoding $\El : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,\Gamma$.
\item
  Inductive function space, with $\Pii : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,(\Gamma\,\ext\,\El\,a) \ra \Ty\,\Gamma$ and application as $\appi :
  \Tm\,\Gamma\,(\Pii\,a\,B)\ra \Tm\,(\Gamma\,\ext\,\El\,a)\,B$ and its
  inverse $\lami$.
\item
  External function space with $\Set_i$ domain: $\Pie : (A : \Set_i)\ra(A \ra
  \Ty\,\Gamma)\ra \Ty\,\Gamma$, with $\appe : \Tm\,\Gamma\,(\Pie\,A\,B)\ra((x :
  A)\ra \Tm\,\Gamma\,(B\,x))$ and its inverse $\lame$.
\item
  Infinitary function space, with $\Piinf : (A : \Set_i)\ra(A \ra
  \Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, $\appinf :
  \Tm\,\Gamma\,(\El\,(\Pie\,A\,b))\ra((x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x)))$
  and $\laminf$.
\item
  An identity type $\Id : (a : \Tm\,\Gamma\,\U)\ra
  \Tm\,\Gamma\,(\El\,a)\ra\Tm\,\Gamma\,(\El\,a)\ra \Tm\,\Gamma\,\U$, with
  $\Refl : (T : \Tm\,\Gamma\,(\El\,a))\ra \Id\,a\,t\,t$, equality reflection
  and UIP.
\end{enumerate}
\end{definition}
In the above listing, we omit most CwF components and equations for substitution and
$\beta\eta$-conversion, but these should be understood to be also part of $\ToS_{i,j}$.

\emph{Notational conventions.} We name elements of $\Con$ as $\Gamma$, $\Delta$,
$\Theta$, elements of $\Sub\,\Gamma\,\Delta$ as $\sigma$, $\delta$, $\nu$,
elements of $\Ty\,\Gamma$ as $A$, $B$, $C$, and elements of $\Tm\,\Gamma\,A$ as
$t$, $u$, $v$. CwF components by default support de Bruijn indices, which are
not easily readable. We use instead a nameful notation for binders in context
extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon\,\ext (a : U) \ext (t :
\El\,a))$. We also define a type-theoretic flavor of $\appi$ for convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[\id, u])\\
  & t\appitt u && :\equiv (\appi\,t)[\id, u]
\end{alignat*}

\begin{definition}[Notion of signature]
A QIIT signature at level $i$ is a context in an arbitrary $M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_i :\equiv (j : \mathbb{N})\ra(M : \ToS_{i,j})\ra Con_M
\]

Here, $i$ refers to the level of external types appearing in the signature, in
the domains of $\Pie$ and $\Piinf$ functions, while the quantified $j$ level is
required to allow interpreting a signature in arbitrary-sized ToS models. Note
that $\Sig_i$ is universe-polymorphic, so it is a family of cETT types and it is
not in any cETT universe.
